Reducing the Memory Cost of Training Convolutional Neural Networks by CPU Offloading

### 2019-07-10 article.tex
Relecture

In recent years, Convolutional Neural Networks (CNN) have enabled 
unprecedented progress on a wide range of computer vision tasks. 
However, training large CNNs is a resource intensive task that requires 
specialized Graphical Processing Units (GPU) and highly optimized 
implementations to get optimal performance from the hardware. 
%Given the ubiquity of CNN for practical computer vision applications,
%considerable efforts have been undertaken to optimize their resource consumption.
GPU memory is a major bottleneck of the CNN training procedure,
 limiting  the size of both inputs and model architectures.
%The backpropagation algorithm requires to accumulate the activation 
%values of hidden layers in live GPU memory during the forward pass
%as these activations are needed for the computation of the weights 
%gradient during the backward pass of the backpropagation algorithm.
In this paper, we propose to alleviate this memory bottleneck by 
leveraging an under-utilized resource of modern systems: 
the device to host bandwidth.
%offloading the input activation of hidden layers to the CPU:
Our method, termed CPU offloading, works by transferring hidden activations to the CPU upon
computation, in order to free GPU memory for upstream layer computations during the forward pass. 
These activations are then transferred back to the GPU as needed by the gradient computations of the backward pass.
The key challenge to our method is to efficiently overlap data transfers and computations
in order to minimize  wall time overhead induced by the additional data transfers.
On a typical work station with a Nvidia Titan X GPU and XXX PCIe lanes, 
we show that our method compares favorably to 
gradient checkpointing as we are able reduce the memory consumption of training a VGG19 model by 35\% with a minimal additional wall time overhead of 21\%. 
Further experiments detail the impact of the different optimization tricks we propose.
Our method is orthogonal to other techniques for memory reduction such as 
quantization and sparsification so that they can easily be combined for further optimizations.

@@

In contrast, our approach does not induce any additional computation:
Instead of computing a set of missing hidden activations during the backward pass,
we propose to offload the hidden activations to the CPU during the forward pass,
and to transfer these activations back to GPU memory during the backward pass.

@@

\begin{equation}
\frac{\delta \mathcal{L}}{\delta \theta_i} = z_{i-1} \star \frac{\delta \mathcal{L}}{\delta z_i} 
\end{equation}

@@

Hence, backpropagation implementations in deep learning frameworks
store hidden layers activations in GPU memory 
upon computation during the forward pass. 
Activations accumulate in live memory buffers throughout the full forward pass
until used for gradients computations in the backward pass. 
Once the gradients computed in the backward pass, 
their associated hidden activation buffers can be freed from live memory. 
However, the accumulation of activation values stored within each layer 
along the forward pass creates a major bottleneck in GPU memory.
In the next Section, we detail our approach to alleviate this memory bottleneck.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=0.8\textwidth]{Figure2.jpg}
\caption{Illustration of the parallel CPU Offloading execution through time.
(top): Computations and data transfer are executed in parallel in their dedicated stream. 
The overhead in wall time $\mathcal{T}_{idle}$ is spent as the computation stream awaits 
for data transfers to complete (illustrated in red). 
(bottom): Without parallelism, computations and data transfers are sequentially executed within the same stream
so that the overhead in wall time corresponds to the total time of data transfers $\mathcal{T}_{data}$}.
\end{figure}

@@

One important exception to this rule concerns skip connections, as illustrated in Figure 3.
Skip connections induce a delay in the GPU buffer de-allocation as the input of residual blocks must
be kept in memory until the end of the residual block computation to be added to the output.
In ResNet architectures, this delay is short enough to have little impact on the memory/wall time trade off.
However, this means that our method is poorly suited to densely connected 
networks such as DenseNet \cite{huang2017densely} or UNet \cite{ronneberger2015u}, which features very long skip connections.

@@

During the backward pass, the input activation $z_{i-1}$ 
of layer $i$ must be transferred back into GPU memory
for the backward gradient computations to proceed.
Hence, we overlap the transfer of $z_{i-1}$ to GPU 
with the backward computations of the upper layer $i+1$ to avoid GPU idle time.
As formalized in Algorithm 2, we thus synchronize the data transfer of $z_{i-1}$ 
with the beginning of the backpropagation through $f_{i+1}$.

@@

We use threading and locks to handle the parallelism and synchronization on the CPU side,
and CUDA streams to handle the parallelization on the GPU end. 
CUDA events provides synchronization primitives to precisdely track 
the completion of kernel executions.
Algorithm 1 and 2 provide pseudo-code for the forward and backward pass respectively.

@@

% Discussion on compute time vs. transfer time.
With the above parallelization scheme, the overhead $\mathcal{T}_{idle}$ in wall time 
is given by the sum of the difference in the computation and data transfer time at each layer.
If data transfer is faster than the computation in each layer, 
CPU offloading will reduce memory consumption without any overhead in wall time.
Hence, CPU offloading is best suited for computation intensive operations like convolution.
In contrast, batch normalization layers come with negligible computational cost,
but significant data transfer time so that offloading the input activations of 
such layer comes with a significant overhead.
Instead of offloading these layers, we reconstruct their input activations 
from their output using their inverse operation, as proposed in \cite{rota2018place},
and only offload the input activations of the convolution and pooling layers to CPU.

@@

\begin{table*}[t]
\begin{center}
\begin{tabular}{ c c }	
\hline
Operation     & Speed (GB/s)  \\
\hline
Pinned Memory Allocation     		&  $1.95$        \\
GPU $\rightarrow$ CPU (Pageable)	&  $1.75$        \\
GPU $\rightarrow$ CPU (Pinned)	        &  $6.5$          \\
CPU $\rightarrow$ GPU (Pageable)	&  $4.1$          \\
CPU $\rightarrow$ GPU (Pinned)	        &  $6.1$          \\
\hline
\end{tabular}
\caption{Measurement of data transfer and pinned memory allocation speed on our workstations.
Pinned memory buffers bring a substantial speed up in both directions of the data transfer.}
\end{center}
\end{table*}

@@

Both parallelization and pinned memory buffers bring substantial improvements 
in speed and have proven critical for performance.
Compression of the hidden states to either sparse representations or lower precision
encodings may bring similar speed-up and would be very interesting to investigate in future work. 	
We expect that simply converting the hidden activations to FP16 before sending them
to the host device should further reduce the data transfer by half.

@@

Searching for an optimal parallelization strategy given a fixed computational graph and hardware system
is a very interesting problem left open to investigate. 
In this paper, we provide a proof of concept by showing that 
even a rather naive parallelization strategies with minimum optimization
leads to a memory/time trade-off competitive relative to 
the well established method of gradient checkpointing.

@@

Convolutional Neural Networks have become the backbone of computer vision systems.
Despite their great success, one major drawback of these models is their intense resource consumption:
Training CNNs needs highly optimized implementations leveraging all possible hardware resources.
In this paper, we proposed to leverage GPU-CPU communication, 
an under-utilized resource in typical CNN training pipelines,
to alleviate the GPU bottleneck in training CNNs. 
Our experiment on a standard single-GPU work station shows a more
effective memory/wall time trade-off than gradient checkpointing.
Our approach is orthogonal to other resource optimization approaches 
such as pruning, quantization so these approaches can easily be combined in future work.

>>>>

In recent years, Convolutional Neural Networks (CNN) have enabled 
unprecedented progress on a wide range of computer vision tasks. 
However, training large CNNs is a resource intensive task that requires 
specialized Graphical Processing Units (GPU) and highly optimized 
implementations to get optimal performance from the hardware. 
%Given the ubiquity of CNN for practical computer vision applications,
%considerable efforts have been undertaken to optimize their resource consumption.
GPU memory is a major bottleneck of the CNN training procedure,
 limiting  the size of both inputs and model architectures.
%The backpropagation algorithm requires to accumulate the activation 
%values of hidden layers in live GPU memory during the forward pass
%as these activations are needed for the computation of the weights 
%gradient during the backward pass of the backpropagation algorithm.
In this paper, we propose to alleviate this memory bottleneck by 
leveraging an under-utilized resource of modern systems: 
the device to host bandwidth.
%offloading the input activation of hidden layers to the CPU:
Our method, termed CPU offloading, works by transferring hidden activations to the CPU upon
computation, in order to free GPU memory for upstream layer computations during the forward pass. 
These activations are then transferred back to the GPU as needed by the gradient computations of the backward pass.
The key challenge to our method is to efficiently overlap data transfers and computations
in order to minimize  wall time overheads induced by the additional data transfers.
On a typical work station with a Nvidia Titan X GPU, 
we show that our method compares favorably to 
gradient checkpointing as we are able reduce the memory consumption of training a VGG19 model by 35\% with a minimal additional wall time overhead of 21\%. 
Further experiments detail the impact of the different optimization tricks we propose.
Our method is orthogonal to other techniques for memory reduction such as 
quantization and sparsification so that they can easily be combined for further optimizations.

@@

In contrast, our approach does not introduce any additional computation:
Instead of computing a set of missing hidden activations during the backward pass,
we propose to offload the hidden activations to the CPU during the forward pass,
and to transfer these activations back to GPU memory during the backward pass.

@@

\begin{equation}
\frac{\delta \mathcal{L}}{\delta \theta_i} = z_{i-1}^F \star \frac{\delta \mathcal{L}}{\delta z_i} 
\end{equation}

@@

Hence, backpropagation implementations in deep learning frameworks
store hidden layers activations in GPU memory 
upon computation during the forward pass. 
Activations accumulate in live memory buffers throughout the full forward pass
until used for gradients computations of the backward pass. 
Once the gradients computed during the backward pass, 
their associated hidden activation buffers can be freed from live memory. 
However, the accumulation of activation values stored within each layer 
along the forward pass creates a major bottleneck in GPU memory.
In the next section, we detail our approach to alleviate this memory bottleneck.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=0.8\textwidth]{Figure2.jpg}
\caption{Illustration of the parallel CPU Offloading execution through time.
(bottom): Without parallelism, computations and data transfers are sequentially executed within the same stream
so that the overhead in wall time corresponds to the total time of data transfers $\mathcal{T}_{data}$.
(top): Computations and data transfer are executed in parallel in their dedicated stream. 
The overhead in wall time $\mathcal{T}_{idle}$ is spent as the computation stream awaits 
for data transfers to complete (illustrated in red). 
}
\end{figure}

@@

One important exception to this rule concerns skip connections, as illustrated in Figure 3.
Skip connections induce a delay in the GPU buffer de-allocation as the input of residual blocks must
be kept in memory until the end of the residual block computation to be added to the output.
In ResNet architectures, this delay is short enough to have little impact on the memory/wall time trade off.
However, this means that our method is poorly suited to densely connected 
architectures such as DenseNet \cite{huang2017densely} or UNet \cite{ronneberger2015u} 
as their long-range skip connections introduce large delays in the GPU buffer de-allocation.

@@

During the backward pass, the input activation $z_{i-1}$ 
of layer $i$ must be transferred back into GPU memory
for the backward gradient computations $\delta f_i$ to proceed.
Hence, we overlap the transfer of $z_{i-1}$ to GPU 
with the backward computations $\delta f_{i+1}$ of the upper layer $i+1$ to avoid GPU idle time.
As formalized in Algorithm 2, we thus synchronize the data transfer of $z_{i-1}$ 
with the beginning of the backpropagation through $f_{i+1}$.

@@

We use threading and locks to handle the parallelism and synchronization on the CPU side,
and CUDA streams to handle the parallelization on the GPU device. 
CUDA events provides synchronization primitives to precisely track 
the completion of kernel executions.
Algorithm 1 and 2 provide pseudo-code for the forward and backward pass respectively.

@@

% Discussion on compute time vs. transfer time.
With the above parallelization scheme, the overhead $\mathcal{T}_{idle}$ in wall time 
is given by the sum of the difference in the computation and data transfer time at each layer.
If data transfer is faster than the computation in each layer, 
CPU offloading will reduce memory consumption without any overhead in wall time.
Hence, CPU offloading is best suited for computation intensive operations like convolutions.
In contrast, batch normalization layers come with negligible computational cost,
but significant data transfer time so that offloading the input activations of 
such layer comes with a significant overhead.
Instead of offloading these layers, we reconstruct their input activations 
from their output using their inverse operation, as proposed in \cite{rota2018place},
and only offload the input activations of the convolution and pooling layers to CPU.

@@

\begin{table*}[t]
\begin{center}
\begin{tabular}{ c c }	
\hline
Operation     & Speed (GB/s)  \\
\hline
Pinned Memory Allocation     		&  $1.95$        \\
GPU $\rightarrow$ CPU (Pageable)	&  $1.75$        \\
GPU $\rightarrow$ CPU (Pinned)	        &  $6.5$          \\
CPU $\rightarrow$ GPU (Pageable)	&  $4.1$          \\
CPU $\rightarrow$ GPU (Pinned)	        &  $6.1$          \\
\hline
\end{tabular}
\caption{Measurement of data transfer and pinned memory allocation speed on our workstation.
Pinned memory buffers bring a substantial speed up in both directions of the data transfer.}
\end{center}
\end{table*}

@@

Both parallelization and pinned memory buffers bring substantial improvements 
in speed and have proven critical for performance.
Compression of the hidden states to either sparse representations or lower precision
encodings may bring similar speed-up and would be very interesting to investigate in future work. 	
We expect that simply converting the hidden activations to FP16 before sending them
to the host device should further reduce the data transfer time by half.

@@

Searching for an optimal parallelization strategy given a fixed computational graph and hardware system
is a very interesting problem left open to investigate. 
In this paper, we provide a proof of concept by showing that 
even a rather naive parallelization strategies with minimum optimization
leads to a competitive memory/time trade-off relatively to 
the well established method of gradient checkpointing.

@@

Convolutional Neural Networks have become the backbone of computer vision systems.
Despite their great success, one major drawback of these models is their intense resource consumption:
Training CNNs needs highly optimized implementations leveraging all possible hardware resources.
In this paper, we proposed to leverage GPU-CPU communication, 
an under-utilized resource in typical CNN training pipelines,
to alleviate the GPU bottleneck in training CNNs. 
Our experiment on a standard single-GPU work station shows a more
effective memory/wall time trade-off than gradient checkpointing.
Our approach is orthogonal to other resource optimization approaches 
such as pruning and quantization so these approaches can easily be combined in future work.
